A Practical Optional Type System for Clojure

### 2012-12-06 dissertation/design-choices.tex
Lyndon corrections + others

Occurrence typing, described by Tobin-Hochstadt and Felleisen~\cite{TF10},
plays a major role in capturing many common Clojure idioms.
One such idiom is the \emph{sequence abstraction}, which
consists of the \lstinline|clojure.core| functions \lstinline|seq|, \lstinline|first|, and \lstinline|rest|.

\lstinline|seq| takes an argument that is either a sequence or \lstinline|nil|,
and returns a non-empty sequence if passed a non-empty sequence, or returns
\lstinline|nil| if passed an empty sequence or \lstinline|nil|.
It is a common idiom to test \lstinline|(seq coll)| in a conditional
to determine whether the sequence \lstinline|coll| is empty:
if \lstinline|(seq coll)| is a true value, then \lstinline|coll|
is non-empty, otherwise it should be considered empty.
Occurrence typing allows us to capture this idiom as shown in
listing \ref{lst:seqtype}.

@@

This function type takes an argument of type \lstinline|(U nil (Seqable x))| (the syntax \lstinline[mathescape]{(U $\overrightarrow{t}$)}
constructs a union of types \lstinline|t|), and returns a type \lstinline|(U nil (Seqable x))|.
Occurrence typing allows us to attach propositions to the return
type of a function. Roughly, these say if \lstinline|(seq coll)|
returns a true value, then \lstinline|coll| is of type
\lstinline|(CountRange 1)| (a type representing a sequence of length 1 or greater),
otherwise \lstinline|coll| is \emph{either} of type \lstinline|nil| or of type \lstinline|(ExactCount 0)| (a type
representing an empty sequence).
(See Section \ref{sec:filterneg} for a more detailed explanation of the filter syntax).

Correspondingly, \lstinline|first| which returns the first member of a sequence,
can be given a more accurate (and surprisingly subtle) type.
\lstinline|first| is overloaded in a similar fashion to \lstinline|seq|.
\lstinline|first| takes an argument that is either a sequence or \lstinline|nil|.
If it is passed \lstinline|nil| or an empty sequence, it returns \lstinline|nil|. Otherwise, if it is passed
non-empty sequence, it returns the first member of the sequence.
Listing \ref{lst:firsttype} shows the type of \lstinline|first|.

@@

The function type has three arities, the last of which can be considered a ``default'' arity.
The first two arities correspond exactly to the two cases described for when 
\lstinline|first| is passed an empty sequence or \lstinline|nil|, and
when \lstinline|first| is passed a non-empty sequence
(the syntax \lstinline[mathescape]{(I $\overrightarrow{t}$)}
constructs an intersection of types \lstinline|t|).

These two functions interact via occurrence typing.
For example, listing \ref{lst:seqocc} tests
\lstinline|(seq coll)| which means occurrences of \lstinline|coll|
in the ``then'' branch are non-empty and those in the
``else'' branch are empty.
We can safely call \lstinline|(.method (first coll))| in the ``then''
branch (assuming elements of \lstinline|coll| have a no-argument method called \lstinline|method|)
always avoiding a \emph{NullPointerException}.

@@

Strickland, Tobin-Hochstadt, and Felleisen describe variable-arity poly\-morphism ~\cite{STF09},
which has been implemented in Typed Racket.
Typed Clojure directly ports most of Typed Racket implementation of variable-arity polymorphism,
as described in Section \ref{sec:impvariablearity}

Some common idiomatic Clojure functions like \lstinline|assoc|, \lstinline|hash-map|,
and \lstinline|partial| cannot be typed by Strickland et al.\ .
Instead, Typed Clojure hard-codes some of these functions as primitives that cannot
be used in a higher-order context.
Expressing the types of these functions is set as future work (Section \ref{sec:futurearity}).

@@

This scheme should be considered experimental, and is not completely fleshed out or implemented.
Notably, the component type \lstinline|c| seems redundant, and Typed Clojure currently
does not provide syntax to ``cast'' an unwriteable array to a be writeable.

@@

This decision, however, needs further work to be completely satisfying.
In particular, type inference should be improved in certain common situations.
For example, we currently do not know how to infer \lstinline|(filter identity coll)|
as returning a sequence of non-\lstinline|nil| elements, for any sequence \lstinline|coll|
(see Section \ref{sec:filterneg}).
Also, Typed Clojure lets the programmer define a set of unenforced \emph{assumptions}
where \emph{null} is allowed in Java fields and methods (see Section \ref{sec:javainterop}),
which is potentially brittle: there is nothing stopping the Java code from introducing unexpected \emph{null} values at runtime.
Adding runtime assertions that enforce these assumptions would allow more
accurate and earlier errors, an approach similar to that taken by Typed Racket~\cite{Tob10} when
interfacing with untyped Racket code.

@@

Protocols and datatypes are two of Clojure's main means of abstraction.
Protocols define a set of unimplemented methods, and each must be assigned types.
Datatype definitions are type checked by ensuring the methods it
implements match the types defined by the Java interface or assigned to the protocol method
it is extending.

>>>>

Occurrence typing, described by Tobin-Hochstadt and Felleisen~\cite{TF10},
plays a major role in capturing many common Clojure idioms.
One such idiom is Clojure's \emph{sequence abstraction}, which
provides a common interface to all collections.
For example, \lstinline|seq| tests whether a collection is empty,
\lstinline|first| returns the first element of a collection,
and \lstinline|rest| returns the tail of its collection argument as a new sequence.
While these summaries describe their most common usages, they do not tell the whole story.

\lstinline|seq|'s description might suggest it is better named \lstinline|not-empty?|,
but \lstinline|seq| is perhaps surprising in two main ways: 
it accepts \lstinline|nil| as an argument (returning \lstinline|nil|),
and, rather than returning \lstinline|true| or \lstinline|false|, returns
a non-empty sequence or \lstinline|nil| respectively.

From a typing perspective, the subtlety of \lstinline|seq| is expressing that \lstinline|(seq coll)|

\begin{itemize}
  \item is a true value when \lstinline|coll| is a non-empty collection, and
  \item is a false value when \lstinline|coll| is either \lstinline|nil| or an empty collection.
\end{itemize}

Occurrence typing allows us to capture this idiom (Listing \ref{lst:seqtype}).

@@

This function type (Listing \ref{lst:seqtype}) 
takes an argument of type \lstinline|(U nil (Seqable x))| (the syntax \lstinline[mathescape]{(U $\overrightarrow{t}$)}
constructs a union of types \lstinline|t|), and returns a type \lstinline|(U nil (Seqable x))|.
Occurrence typing allows us to attach propositions to the return
type of a function. Roughly, these say if \lstinline|(seq coll)|
returns a true value, then \lstinline|coll| is of type
\lstinline|(CountRange 1)| (a type representing a sequence of length 1 or greater),
otherwise \lstinline|coll| is \emph{either} of type \lstinline|nil| or of type \lstinline|(ExactCount 0)| (a type
representing an empty sequence).
(See Section \ref{sec:filterneg} for a more detailed explanation of the filter syntax).

\lstinline|first| is overloaded in a similar fashion to \lstinline|seq|:
\lstinline|(first coll)| returns

\begin{itemize}
  \item \lstinline|nil| if \lstinline|coll| is \lstinline|nil| or an empty collection, and
  \item returns the first item of the collection if \lstinline|coll| is a non-empty collection.
\end{itemize}

Listing \ref{lst:firsttype} shows the type of \lstinline|first|.
The main thing to grasp from this definition is that three function types are given:
the first type handling the first case above, the second type handling the second case above,
and the third arity acting as a base case, general enough to catch both cases.

@@

\lstinline|seq| and \lstinline|first| interact via occurrence typing.
For example, Listing \ref{lst:seqocc} tests
\lstinline|(seq coll)| which means occurrences of \lstinline|coll|
in the ``then'' branch are non-empty and those in the
``else'' branch are empty.
We can safely call \lstinline|(.method (first coll))| in the ``then''
branch (assuming elements of \lstinline|coll| have a no-argument method called \lstinline|method|)
always avoiding a \emph{NullPointerException}.

@@

Strickland, Tobin-Hochstadt, and Felleisen describe variable-arity poly\-morphism ~\cite{STF09},
which has been implemented in Typed Racket.
Typed Clojure directly ports most of Typed Racket's implementation of variable-arity polymorphism,
as described in Section \ref{sec:impvariablearity}.

Some common idiomatic Clojure functions like \lstinline|assoc|, \lstinline|hash-map|,
and \lstinline|partial| cannot be typed by Strickland \emph{et al.}.
Instead, Typed Clojure hard-codes some of these functions as primitives that cannot
be used in a higher-order context.
Expressing the types of these functions is set as future work (Section \ref{sec:futurearity}).

@@

This scheme should be considered experimental, and is not completely fleshed out or implemented.
Notably, the component type \lstinline|c| seems redundant, and Typed Clojure currently
does not provide syntax to ``cast'' an unwriteable array to be writeable.

@@

This decision, however, needs further work to be completely satisfying.
In particular, type inference should be improved in certain common situations.
For example, we currently do not know how to infer \lstinline|(filter identity coll)|
as returning a sequence with the elements of \lstinline|coll| that are not \lstinline|nil| or \lstinline|false|,
for any sequence \lstinline|coll|
(see Section \ref{sec:filterneg}).
Also, Typed Clojure lets the programmer define a set of unenforced \emph{assumptions}
where \emph{null} is allowed in Java fields and methods (see Section \ref{sec:javainterop}),
which is potentially brittle: there is nothing stopping the Java code from introducing unexpected \emph{null} values at runtime.
Adding runtime assertions that enforce these assumptions would allow more
accurate and earlier errors, an approach similar to that taken by Typed Racket~\cite{Tob10} when
interfacing with untyped Racket code.

@@

Protocols and datatypes are two of Clojure's main means of abstraction.
Protocols define a set of unimplemented methods, and each must be assigned types.
Datatype definitions are type checked by ensuring the methods they
implement match the types defined by the Java interface or assigned to the protocol method
they are extending.

### 2012-12-06 dissertation/experiments.tex
Lyndon corrections + others

This section describes porting a function from \emph{clojure.contrib.reflect}
\footnote{\emph{clojure.contrib} Github project: https://github.com/richhickey/clojure-contrib},
a Clojure library that relies heavily on Java interoperability.
I chose to port \lstinline|call-method| for several reasons: it chains several Java calls together,
it uses primitive arrays, and \lstinline|null| is a valid value in one place.

@@

The final successfully type checked code is presented in Listing \ref{lst:callmethodfinal}.

\begin{lstlisting}[caption=Type Annotated code for call-method, label=lst:callmethodfinal]
(non-nil-return java.lang.Class/getDeclaredMethod :all)
(nilable-param java.lang.reflect.Method/invoke {2 #{0}})

@@

This experiment involved porting an implementation of red-black trees
used as an experiment for SML CIDRE by Davies~\cite{Dav05}.
SML CIDRE is a sort-checker for Standard ML that supports refinement
types.
The red-black tree implementation was ported to work with SML CIDRE
to statically check red-black tree invariants.
It is a particularly good experiment for SML CIDRE because it generates an unusually large
amount of intersection types.

Intersection types in SML CIDRE are heavily optimised via memoisation.
Because of this, SML CIDRE is able to check this experiment with little trouble.
The current Typed Clojure prototype is not as successful, and appears to hang
during type checking.
The reasons are not entirely clear, but it is likely the number of intersection types
are too large for the current prototype to handle, which has no optimisations
around intersection types.

This experiment was intended to stress test Typed Clojure to an extreme, and
it is unsurprising that Typed Clojure was not up to the task.
One interesting point was noticed when porting it.
The SML CIDRE version of the red-black tree experiment
uses SML CIDRE datasorts to represent the red-black tree invariants.
The Typed Clojure version entirely uses plain hash-maps, a common Clojure idiom.
It was noticed that the heterogeneous map types that Typed Clojure provides would be sufficient to
represent the red-black tree invariants.

@@

\begin{lstlisting}[caption=Several monad types, label=lst:severalmonads]
; Maybe monad
(ann maybe-m (MonadPlusZero
               (TFn [[x :variance :covariant]] 
                 (U nil x))))

; Sequence monad (called "list monad" in Haskell)
(ann sequence-m (MonadPlusZero 
                  (TFn [[x :variance :covariant]]
                    (Seqable x))))

; State monad
(def-alias State
  (TFn [[r :variance :covariant]
        [s :variance :invariant]]
    [s -> '[r s]]))
(ann state-m (All [s]
               (Monad (TFn [[x :variance :covariant]]
                        (State x s)))))
\end{lstlisting}

@@

Listing \ref{lst:maybet} shows the full code for type checking the definition
of the maybe monad transformer.
Typing this example was a matter of assigning types to anonymous functions
(similar to standard monad definitions) and manually instantiating 
polymorphic types with \lstinline|inst|
that parameterise over higher kinds, which cannot be inferred by Typed Clojure.

\begin{lstlisting}[caption={Maybe monad transformer in Typed Clojure}, label=lst:maybet]
(ann maybe-t
     (All [[m :kind (TFn [[x :variance :covariant]] Any)]]
       (Fn 
         [(AnyMonad m) -> (MonadPlusZero (TFn [[y :variance :covariant]]
                                           (m (U nil y))))]
         [(AnyMonad m) nil -> (MonadPlusZero (TFn [[y :variance :covariant]]
                                               (m (U nil y))))]
         [(AnyMonad m) nil (U ':m-plus-default ':m-plus-from-base)
          -> (MonadPlusZero (TFn [[y :variance :covariant]]
                              (m (U nil y))))])))
(defn maybe-t
  "Monad transformer that transforms a monad m into a monad in which
   the base values can be invalid (represented by nothing, which defaults
   to nil). The third argument chooses if m-zero and m-plus are inherited
   from the base monad (use :m-plus-from-base) or adopt maybe-like
   behaviour (use :m-plus-from-transformer). The default is :m-plus-from-base
   if the base monad m has a definition for m-plus, and
   :m-plus-from-transformer otherwise."
  ([m] ((inst maybe-t m) m nil :m-plus-default))
  ([m nothing] ((inst maybe-t m) m nothing :m-plus-default))
  ([m nothing which-m-plus]
   (monad-transformer m which-m-plus
     [m-result (with-monad m m-result)
      m-bind   (with-monad m
                 (ann-form
                   (fn m-bind-maybe-t [mv f]
                     (m-bind
                       mv
                       (ann-form
                         (fn [x]
                           (if (nil? x)
                             (m-result nothing) 
                             (f x)))
                         [(U nil a) -> (m (U nil b))])))
                   (All [a b]
                     [(m (U nil a)) [a -> (m (U nil b))] -> (m (U nil b))])))
      m-zero   (with-monad m (m-result nothing))
      m-plus   (with-monad m
                 (ann-form
                   (fn m-plus-maybe-t [& mvs]
                     (if (empty? mvs)
                       (m-result nothing)
                       (m-bind (first mvs)
                               (ann-form
                                 (fn [v]
                                   (if (= v nothing)
                                     (apply m-plus-maybe-t (rest mvs))
                                     (m-result v)))
                                 [(U x nil) -> (m (U x nil))]))))
                   (All [x] [(m (U x nil)) * -> (m (U x nil))])))
      ])))

\end{lstlisting}

@@

Listing \ref{lst:conduittypes} shows the type constructor for conduit,
\lstinline|==>|. Notice the body of \lstinline|==>| includes a reference
to itself in a position where variance must be known in advance.
\lstinline|declare-alias-kind| is used to declare the kind of an alias.
Once an alias with an already declared kind is defined with \lstinline|def-alias|,
the defined kind must match the declared kind, otherwise a type error is thrown.
(The specifics of the types in Listing \ref{lst:conduittypes} are not relevant to the rest of this discussion,
so they are not explained).

\begin{lstlisting}[caption=Types for Conduits, label=lst:conduittypes]
...

(def-alias Result
  (TFn [[x :variance :covariant]]
    (U nil ;stream is closed
       '[] ;abort/skip
       '[x];consume/continue
       )))

@@

(def-alias ==>
  (TFn [[in :variance :contravariant] 
        [out :variance :invariant]]
    [in -> '[(U nil (==> in out)) (Cont out out)]]))

...
\end{lstlisting}

>>>>

This section describes porting a function from \emph{clojure.contrib.reflect}
\footnote{\emph{clojure.contrib} Github project: https://github.com/richhickey/clojure-contrib}
\footnote{Note that while the \emph{clojure.contrib} library as a whole should be considered deprecated, useful
code still exists within it which apparently has not been migrated elsewhere.},
a Clojure library that relies heavily on Java interoperability.
I chose to port \lstinline|call-method| for several reasons: it chains several Java calls together,
it uses primitive arrays, and \lstinline|null| is a valid value in one place.

@@

The final successfully type checked code is presented in Listing \ref{lst:callmethodfinal}.
\\\\
% bump to next page
\begin{lstlisting}[caption=Type Annotated code for call-method, label=lst:callmethodfinal]
(non-nil-return java.lang.Class/getDeclaredMethod :all)
(nilable-param java.lang.reflect.Method/invoke {2 #{0}})

@@

This experiment involved porting an implementation of red-black trees
used as an experiment for SML CIDRE by Davies~\cite{Dav05}.
SML CIDRE is a sort-checker for Standard ML that supports refinement
types.
The red-black tree implementation was ported to work with SML CIDRE
to statically check red-black tree invariants.
It is a particularly good experiment for SML CIDRE because it generates an unusually large
number of intersection types.

Intersection types in SML CIDRE are heavily optimised via memoisation.
Because of this, SML CIDRE is able to check this experiment with little trouble.
The current Typed Clojure prototype is not as successful, and appears to hang
during type checking. As this experiment was particularly ambitious,
diagnosing and fixing this issue is delegated to future work.

One interesting point was noticed when porting the experiment.
The SML CIDRE version of the red-black tree experiment
uses SML CIDRE datasorts to represent the red-black tree invariants.
The Typed Clojure version entirely uses plain hash-maps, a common Clojure idiom.
It was noticed that the heterogeneous map types that Typed Clojure provides would be sufficient to
represent the red-black tree invariants.

@@

\begin{lstlisting}[caption=Several monad types, label=lst:severalmonads]
; Maybe monad
(ann maybe-m (MonadPlusZero
               (TFn [[x :variance :covariant]] 
                 (U nil x))))
; Sequence monad (called "list monad" in Haskell)
(ann sequence-m (MonadPlusZero 
                  (TFn [[x :variance :covariant]]
                    (Seqable x))))
; State monad
(def-alias State
  (TFn [[r :variance :covariant]
        [s :variance :invariant]]
    [s -> '[r s]]))
(ann state-m (All [s]
               (Monad (TFn [[x :variance :covariant]]
                        (State x s)))))
\end{lstlisting}

@@

Finally, Listing \ref{lst:maybet} shows the type assigned to the maybe monad transformer definition.
The most interesting aspect to notice here is that the monad \lstinline|m| is kept abstract throughout
the type.

\begin{lstlisting}[caption={Maybe monad transformer type}, label=lst:maybet]
(ann maybe-t
     (All [[m :kind (TFn [[x :variance :covariant]] Any)]]
       (Fn 
         [(AnyMonad m) -> (MonadPlusZero (TFn [[y :variance :covariant]]
                                           (m (U nil y))))]
         [(AnyMonad m) nil -> (MonadPlusZero (TFn [[y :variance :covariant]]
                                               (m (U nil y))))]
         [(AnyMonad m) nil (U ':m-plus-default ':m-plus-from-base)
          -> (MonadPlusZero (TFn [[y :variance :covariant]]
                              (m (U nil y))))])))
\end{lstlisting}

@@

Listing \ref{lst:conduittypes} shows the type constructor for a conduit type,
\lstinline|==>|. Notice the body of \lstinline|==>| includes a reference
to itself in a position where variance must be known in advance.
\lstinline|declare-alias-kind| is used to declare the kind of an alias.
Once an alias with an already declared kind is defined with \lstinline|def-alias|,
the defined kind must match the declared kind, otherwise a type error is thrown.
(The specifics of the types in Listing \ref{lst:conduittypes} are not relevant to the rest of this discussion,
so they are not explained).

\begin{lstlisting}[caption=Types for Conduits, label=lst:conduittypes]
(def-alias Result
  (TFn [[x :variance :covariant]]
    (U nil ;stream is closed
       '[] ;abort/skip
       '[x];consume/continue
       )))

@@

(def-alias ==>
  (TFn [[in :variance :contravariant] 
        [out :variance :invariant]]
    [in -> '[(U nil (==> in out)) (Cont out out)]]))
\end{lstlisting}

### 2012-12-06 dissertation/implementation.tex
Lyndon corrections + others

\begin{itemize}
  \item Pairwise, each \lstinline|S| is a below \lstinline|T|,
  \item \lstinline|R| is below \lstinline|expected|, if \lstinline|expected| is provided.
\end{itemize}

@@

This returns a substitution that replaces occurrences of \lstinline|x| with \lstinline|true|
and \lstinline|y| with \lstinline|Any|.
This helps infer a result type of \lstinline|(constantly true)|
as \lstinline|[Any * -> true]|.

@@

Porting Typed Racket's variable-arity polymorphism implementation was
tedious because some of the relevant internal functions interact
in strange ways with the rest of Typed Racket. My reaction was that Typed Racket
was initially designed without variable-arity polymorphism and was added
without major changes to other components. Typed Clojure was
developed with the same design so full variable-arity polymorphism
implementation could be ported without change.

@@

To better understand the problem, listing \ref{lst:filterann} presents the type of \lstinline|filter|,
which takes a function \lstinline|f| and a sequence \lstinline|s| as arguments, and returns a sequence that 
contains each element in \lstinline|s| such that applying \lstinline|f| to the element returns a true value.
The \lstinline|:filters| syntax requires some explanation. 
Function types support an optional \emph{filter set} attached to its return type, written as a map
with \lstinline|:then| and/or \lstinline|:else| keys (if omitted, they default to the trivially true proposition which has no effect). 
The ``then-proposition'' and ``else-proposition'' are added
to the type environment when the return value is a true and false value respectively.
For example, the filter set \lstinline|{:then (is y 0)}| is read ``if the return value is a true value, then the first argument
must be of type y, otherwise if it is a false value, nothing interesting is enforced''
The type given for \lstinline|filter| works because the type variable
\lstinline|y| occurs in both the ``then-proposition'' of the first argument and the return type 
\lstinline|(Seqable y)|.

\begin{lstlisting}[caption=Troublesome filter, label=lst:filtertrouble]
(filter (fn [a] (not (nil? a))) coll)|
\end{lstlisting}

The difficulty starts with something like listing \ref{lst:filtertrouble},
where the inferred filter set for the first argument to \lstinline|filter|
is \lstinline|{:then (! nil 0) :else (is nil 0)}|\footnote{(! nil 0) is the proposition that the first argument is \emph{not}
of type \lstinline|nil|.}. The ``then-proposition'' \lstinline|(! nil 0)| does not fit with
\lstinline|(is y 0)| that \lstinline|filter| expects.

We can sometimes get around this if we already have a predicate with a positive
``then-proposition''. For example, if we are filtering out \lstinline|nil| values
from a sequence of type \lstinline|(Seqable (U Number nil))|, we can replace
listing \ref{lst:filtertrouble} with \lstinline|(filter number? coll)|,
where \lstinline|number?| has the filter set \lstinline|{:then (is Number 0) :else (! Number 0)}|.
This does not work, however, when filtering a sequence of type like \lstinline|(Seqable (U x nil))|
for some unspecified \lstinline|x| because there is no built-in predicate with ``then-proposition''
\lstinline|(is x 0)|.

@@

Using this technique, the anonymous function in listing 
\ref{lst:filtergood}\footnote{The Clojure syntax \lstinline|#(not (nil? \%))| 
is equivalent to \lstinline|(fn [a] (not (nil? a)))|}
has the filter set \lstinline|{:then (& (! nil 0) (is x 0)) :else (is nil 0)}|
which is good enough to infer the \lstinline|filter|ed result as \lstinline|(Seqable x)|.

>>>>

\begin{itemize}
  \item Pairwise, each \lstinline|S| is a subtype of \lstinline|T|,
  \item \lstinline|R| is below \lstinline|expected|, if \lstinline|expected| is provided.
\end{itemize}

@@

This returns a substitution that replaces occurrences of \lstinline|x| with \lstinline|true|
and \lstinline|y| with \lstinline|Any|.
This helps infer the type of the expression \lstinline|(constantly true)|
as \lstinline|[Any * -> true]| (where \lstinline|true| is the singleton type containing just the value \lstinline|true|).

@@

Porting Typed Racket's variable-arity polymorphism implementation was
tedious because some of the relevant internal functions interact
in strange ways with the rest of Typed Racket. My impression was that Typed Racket
was initially designed without variable-arity polymorphism and was added
without major changes to other components. Typed Clojure was
developed with the same design so full variable-arity polymorphism
implementation could be ported without change.

@@

To better understand the problem, Listing \ref{lst:filterann} presents the type of \lstinline|filter|,
which takes a function \lstinline|f| and a sequence \lstinline|s| as arguments, and returns a sequence that 
contains each element in \lstinline|s| such that applying \lstinline|f| to the element returns a true value.
The \lstinline|:filters| syntax requires some explanation. 
Function types support an optional \emph{filter set} attached to the return type, written as a map
with \lstinline|:then| and/or \lstinline|:else| keys (if omitted, they default to the trivially true proposition which has no effect). 
The ``then-proposition'' and ``else-proposition'' are added
to the type environment when the return value is a true and false value respectively.
For example, the filter set \lstinline|{:then (is y 0)}| is read ``if the return value is a true value, then the first argument
must be of type y, otherwise if it is a false value, nothing interesting is enforced''
The type given for \lstinline|filter| works because the type variable
\lstinline|y| occurs in both the ``then-proposition'' of the first argument and the return type 
\lstinline|(Seqable y)|.

\begin{lstlisting}[caption=Troublesome filter, label=lst:filtertrouble]
(filter (fn [a] (not (nil? a))) coll)
\end{lstlisting}

The difficulty starts with something like Listing \ref{lst:filtertrouble},
where the inferred filter set for the first argument to \lstinline|filter|
is \lstinline|{:then (! nil 0) :else (is nil 0)}|\footnote{(! nil 0) is the proposition that the first argument is \emph{not}
of type \lstinline|nil|.}. The ``then-proposition'' \lstinline|(! nil 0)| does not fit with
\lstinline|(is y 0)| that \lstinline|filter| expects.

We can sometimes get around this if we already have a predicate with a positive
``then-proposition''. For example, if we are filtering out \lstinline|nil| values
from a sequence of type \lstinline|(Seqable (U Number nil))|, we can replace
Listing \ref{lst:filtertrouble} with \lstinline|(filter number? coll)|,
where \lstinline|number?| has the filter set \lstinline|{:then (is Number 0) :else (! Number 0)}|.
This does not work, however, when filtering a sequence of type like \lstinline|(Seqable (U x nil))|
for some unspecified \lstinline|x| because there is no built-in predicate with ``then-proposition''
\lstinline|(is x 0)|.

@@

Using this technique, the anonymous function in Listing 
\ref{lst:filtergood}\footnote{The Clojure syntax \lstinline|#(not (nil? \%))| 
is equivalent to \lstinline|(fn [a] (not (nil? a)))|}
has the filter set \lstinline|{:then (& (! nil 0) (is x 0)) :else (is nil 0)}|
which is good enough to infer the \lstinline|filter|ed result as \lstinline|(Seqable x)|.
